Supplemental Material: Multidimensional analysis and detection of informative features in human white matter

### 2021-01-21 reports/manuscript/sections/results.tex
ENH: update paper and figures

\section*{Results and discussion}

We developed a method for analyzing dMRI tractometry data with SGL. We
demonstrate the use of this method on two previously published datasets in both
a classification setting and a regression setting.

\subsection*{SGL accurately detects ALS in tractometry data in a classification setting}

Using data from a previous study of the corticospinal tract (CST) profile and
ALS\cite{sarica2017corticospinal}, we tested the performance of SGL in a
classification setting. The previous study predicted ALS status with a mean
accuracy of 80\% using a random forest algorithm based on a priori selection of
features within the corticospinal tract. SGL delivers competitive predictive
performance (mean 93\% $\pm$ 2\% accuracy, 0.978 $\pm$ 0.006 ROC AUC) without
the need for a priori feature engineering. The results of the classification
prediction are shown in Figure \ref{fig:class-results} with ``ground-truth'' ALS
status separated into columns, and predicted ALS status encoded by color. In
addition to this classification performance, SGL also identifies the white
matter tracts most important for ALS classification. The relative importance of
white matter features is captured in the $\beta$ coefficients from
Eq~\eqref{eq:sgl}. Figure \ref{fig:class-profiles} depicts these coefficients along
the right CST, plotted over the FA values for the control and ALS subject
groups. We find that SGL selects FA metrics in the corticospinal tract and
particularly in the right corticospinal tract as most important to ALS
classification, confirming previous findings\cite{van2011upper,
toosy2003diffusion, sarica2014tractography, sage2007quantitative,
sage2009quantitative, karlsborg2004corticospinal, ellis1999diffusion,
cosottini2005diffusion, ciccarelli2009investigation, abe2010voxel} and
identifying the portions of the brain that were selected \emph{a priori} in the
previous study from which we collected our data\cite{sarica2017corticospinal}.

Analyzing the ways in which the model mislabels individuals may also provide
insight. We found that mislabelled subjects are outliers relative to their group
with respect to diffusion features of the CST. Figure \ref{fig:class-errors}
depicts the group FA values along with FA values of mislabelled subjects, two
false negatives and one false-positive. The false negative classifications have
high FA in one of the two sections of the CST where $||\hat{\beta}|| > 0$ in
Figure \ref{fig:class-profiles}. The false positive subject has an FA that
oscillates between the two group means. Thus, the SGL method fails
comprehensibly.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.45\textwidth]{classification_probs.png}
    \includegraphics[width=0.45\textwidth]{classification_beta_bupu.png}

    \caption{{\bf SGL accurately predicts ALS.}
        Left: Classification probabilities for each subject's ALS diagnosis.
        Controls are on the left while patients are on the right. Predicted
        controls are in blue and predicted patients are in red. Thus, false
        positive are represented as red dots on the left, while false negatives
        are represented as blue dots on the right. The SGL algorithm achieves
        93\% $\pm$ 2\% accuracy,with 0.978 $\pm$ 0.006 ROC AUC. Right: SGL
        coefficients are presented on a skeleton of the major tracts. The brain
        is oriented with the right hemisphere to our left and anterior out of
        the page. As expected large negative coefficients are in the FA of the
        CST (and particularly in the right hemisphere, here to the left)}
    \label{fig:class-results}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.95\textwidth]{classification_tract_profiles.pdf}
    \caption{{\bf Model coefficients mirror FA differences}. The places along the
       length of the CST  where $\hat{\beta}$ coefficients for FA
       (dashed line, right axis) have large negative values correspond to the
       locations of substantial differences between the ALS (green) and control
       (orange) FA (shaded area indicates standard error of the mean).}
    \label{fig:class-profiles}
\end{figure}

\begin{figure}[h]
    \centering
        \includegraphics[width=0.95\textwidth]{classification_subjects_profiles.pdf}
    \caption{{\bf Model mis-classifications correspond to features identified by
       the model}. The FA in CST of individuals that are mis-classified by the
       model is compared the group FA (with shaded are indicating standard error
       of the mean). False negative classifications (individuals that have ALS,
       but are classified as patients) correspond to high FA either in one of
       the two regions of large $||\hat{\beta}||$ in
       Figure \ref{fig:class-profiles}. The false positive classification has
       an FA profile that oscillates between the two group means.
   }
    \label{fig:class-errors}
\end{figure}

\subsection*{SGL accurately predicts age from tractometry data in a regression setting}

To test the performance of SGL with tractometry data in a continuous regression
task, we focus here on the prediction of biological age based on tractometry
data. Prediction of ``brain age'' is a commonly undertaken task. This is both
because it operates on a natural scale, with meaningful and easily understood
units, as well as because predictions of brain age, and deviations from accurate
prediction are diagnostic of overall brain health (for a recent review, see
\cite{Cole2019-rz}). The WH dataset used here contains data from 76 healthy
subjects, ranging between 6 years and 50 years of age
\cite{yeatman2014lifespan}. In this case, biological age was used as the
predicted variable ($y$ in equation \ref{eq:lm}). SGL was fit to
tractometry-extracted features: FA and MD in 20 major brain tracts, with each
tract divided into 100 nodes. To evaluate the fit of the model, we used a nested
cross-validation procedure. In this procedure, batches of subjects are held out.
For each batch (or fold), the model is fully fit without this data. Then, once
the parameters are fixed, the model is inverted to predict the ages of held out
subjects based on the linear coeffiecients and the static non-linearity. This
scheme automatically finds the right level of regularization (i.e., sparseness)
and fits the coefficients to the ill-posed linear model, while guarding against
overfitting. SGL accurately predicts the age of the subjects in this procedure,
with a mean absolute error of 3.6 years (Figure \ref{fig:regress-results}, left
panel). This is lower than the results of a recent study that predicted age in a
large sample, based on diffusion MRI features \cite{Richard2018-ux}.
Nevertheless, older subjects have higher residual variance, reflecting the
automatically-chosen log-transformation and implying that brain age becomes more
difficult to predict as we age chronologically (\ref{fig:regress-results}, right
panel). The model weights are distributed over many different tracts and dMRI
tissue properties (Figure \ref{fig:regress-beta} left). This demonstrates that
SGL is not coerced to produce overly sparse results when a more accurate model
requires a dense selection of features. Furthermore, looking closer at a
selection of tracts where high coefficients are found demonstrates that
diffusion properties (FA, in this case) are different in different age groups in
parts of the tracts where these higher coefficients are found (Figure
\ref{fig:regress-beta} right).

\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{regression_residuals.png}
    \caption{{\bf Predicting age with tractometry and SGL.} Left: The predicted
    age of each individual (on the abscissa) and true age (on the ordinate),
    from the test splits (i.e., when each subject's data was held out in fitting
    the model); an accurate prediction falls close to the $y=x$ line (dashed).
    The mean absolute error in this case is 3.6 years and, the coefficient of
    determination $R^2=0.3$. Right: Standardized residuals (on the abscissa) as
    a function of the true age (on the ordinate). Predictions are generally more
    accurate for younger individuals.}
    \label{fig:regress-results}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.3\textwidth]{regression_beta_annotated.pdf}
    \includegraphics[width=0.67\textwidth]{regression_tract_profiles.pdf}
    \caption{{\bf Feature importance for predicting age from tractometry.} Left:
    A skeletonized display of the main brain tracts analyzed, with anterior
    facing up, and right hemisphere on the right. The $\hat{\beta}$ coefficients
    displayed in blue (negative) to red (positive) are for measurements of FA
    along the length of the tracts. The left cingulum cingulate (A) and forceps
    minor (B) are highlighted. Right: the FA (in shades of blue and green) and
    the $beta$ coefficients (dashed) in (A) left cingulum and (B) forceps
    minor.}
    \label{fig:regress-beta}
\end{figure}

>>>>

\subsection*{Results}

\begin{figure*}
    \vspace{3cm}
    \rlap{
        \includegraphics[width=0.33\textwidth]{als_coefs.pdf}
    }
    \vspace{-8.4cm}
    \includegraphics[width=\textwidth]{als_results.pdf}
    {\phantomsubcaption\label{fig:class-results:probs}}
    {\phantomsubcaption\label{fig:class-results:coefs3d}}
    {\phantomsubcaption\label{fig:class-results:tract-profiles}}
    {\phantomsubcaption\label{fig:class-results:false-positives}}
    {\phantomsubcaption\label{fig:class-results:false-negatives}}
    \caption{%
        {\bf SGL accurately and interpretably predicts ALS diagnosis.}
        \label{fig:class-results}
        {\bf (a)} Classification probabilities for ALS diagnosis, with
        controls on the left, patients on the right, predicted controls in
        blue, and predicted patients in orange. So orange dots on the left
        represent false positives, while blue dots on the right represent
        false negatives. We achieve {\protect\alsAccuracy}\% accuracy with an
        ROC AUC of {\protect\alsRocAuc}.
        {\bf (b)} SGL coefficients are presented on the core fibers of major
        fiber bundles. They exhibit high group sparsity and are concentrated
        in the fa of the CST. The brain is oriented with the right hemisphere
        in the foreground and anterior to the right of the page. The CST\_L,
        CST\_R, and CF\_A bundles are indicated for orientation.
        {\bf (c)} SGL identifies three portions of the CST as important,
        where $\hat{\beta}$ (dashed line, right axis) has large values. These
        are centered around nodes 30, 65, and 90, corresponding to locations
        of substantial differences in FA between the ALS and control groups
        (shaded areas indicates standard error of the mean).
        {\bf (d)} Bundle profiles for false positive classifications. Line
        colors correspond to the marker edge color in the top left plot.
        These individuals have reduced FA in the CST portions which SGL
        identified as important. Their misclassification is coherent with the
        feature importance and the group differences in FA.
        {\bf (e)} Individual bundle profiles for false negative
        classifications. These individuals have bundle profiles which
        oscillate between the group means.
    }
\end{figure*}

We developed a method for analyzing dMRI tractometry data with SGL. We
demonstrate the use of this method on four previously published datasets in
both a classification setting and a regression setting.

\subsubsection*{SGL accurately detects ALS from tractometry data}

Using data from a previous study of the corticospinal tract (CST) profile and
ALS \cite{sarica2017corticospinal}, we tested the performance of SGL in a
classification setting. The previous study predicted ALS status with a mean
accuracy of 80\% using a random forest algorithm based on a priori selection
of features within the corticospinal tract. SGL delivers competitive
predictive performance (mean {\alsAccuracy}\% accuracy, {\alsRocAuc} ROC AUC)
without the need for a priori feature engineering. The results of the
classification prediction are shown in \cref{fig:class-results:probs} with
``ground-truth'' ALS status separated into columns, and predicted ALS status
encoded by color. In addition to this classification performance, SGL also
identifies the white matter tracts most important for ALS classification. The
relative importance of white matter features is captured in the $\beta$
coefficients from \cref{supp:eq:sgl}. \Cref{fig:class-results:tract-profiles}
depicts these coefficients along the right CST, plotted over the FA values
for the control and ALS subject groups. We find that SGL selects FA metrics
in the corticospinal tract and particularly in the right corticospinal tract
as most important to ALS classification, confirming previous findings
\cite{van2011upper, toosy2003diffusion, sarica2014tractography,
sage2007quantitative, sage2009quantitative, karlsborg2004corticospinal,
ellis1999diffusion, cosottini2005diffusion, ciccarelli2009investigation,
abe2010voxel} and identifying the portions of the brain that were selected
\emph{a priori} in the previous study from which we collected our data
\cite{sarica2017corticospinal}.

The $\beta$ coefficients exhibit high bundle level sparsity; only some
bundles are important, which can be be confirmed by observing the value of
$\alpha$ in \cref{supp:eq:sgl}, the regularization hyperparameter that
controls the mixture of the group lasso and lasso penalties. If $\alpha$ is
closer to zero, it indicates that the phenotype in question preferentially
correlates with only a few groups of covariates. For the ALS dataset, $\alpha
= $ \alsLRatio, confirming that the white matter correlates of ALS reside
mostly in one bundle, namely the CST.

Analyzing the ways in which the model mislabels individuals also provides
insight. We found that mislabelled subjects are outliers relative to their
group with respect to diffusion features of the CST.
\Cref{fig:class-results:false-positives,fig:class-results:false-negatives}
depicts the group FA values along with FA values of mislabelled subjects,
four false negatives and three false-positives. The false positive
classifications have reduced FA in one or more of the three sections of the
CST where $\norm*{\hat{\beta}}$ is large
\cref{fig:class-results:tract-profiles}. The false negative subjects have FA
profiles that oscillate between the two group means. Thus, the SGL method
fails comprehensibly.

\begin{figure*}
    \vspace{3.65cm}
    \rlap{
        \hspace{1cm}
        \includegraphics[width=0.275\textwidth]{wh_coefs.pdf}
        \hspace{0.25em}
        \includegraphics[width=0.275\textwidth]{hbn_coefs.pdf}
        \hspace{0.25em}
        \includegraphics[width=0.275\textwidth]{camcan_coefs.pdf}
    }
    \vspace{-7.3cm}
    \includegraphics[width=\textwidth]{regression_scatter.pdf}
    {\phantomsubcaption\label{fig:age-results:wh-scatter}}
    {\phantomsubcaption\label{fig:age-results:hbn-scatter}}
    {\phantomsubcaption\label{fig:age-results:cc-scatter}}
    {\phantomsubcaption\label{fig:age-results:wh-coefs3d}}
    {\phantomsubcaption\label{fig:age-results:hbn-coefs3d}}
    {\phantomsubcaption\label{fig:age-results:cc-coefs3d}}
    {\phantomsubcaption\label{fig:age-results:wh-profile}}
    {\phantomsubcaption\label{fig:age-results:hbn-profile}}
    {\phantomsubcaption\label{fig:age-results:cc-profile}}
    \caption{%
        {\bf Predicting age with tractometry and SGL.}
        \label{fig:age-results}
        {\bf (top)} The predicted age vs. true age of each individual from the test
        splits (i.e., when each subject's data was held out in fitting the
        model) for the {\bf (a)} WH, {\bf (b)} HBN, and {\bf (c)} Cam-CAN
        datasets; an accurate prediction falls close to the $y=x$ line
        (dashed). The mean absolute error (MAE) and coefficient of
        determination $R^2$ are presented in the lower right of each scatter
        plot.
        {\bf (middle)} Feature importance for predicting age from tractometry in
        the {\bf (d)} WH, {\bf (e)} HBN, and {\bf (f)} Cam-CAN datasets.
        The orientation of the
        brain is that same as in \cref{fig:class-results:coefs3d}, however because
        the coefficients exhibit high global sparsity (as opposed to group
        sparsity), we plot the mean of the absolute value of $\hat{\beta}$
        for each bundle on the core fiber. The global distrubution of the
        $\hat{\beta}$ coefficients reflects the fact that aging is not
        confined to a single white matter bundle.
        {\bf (bottom)} Age quintile bundle profiles for the {\bf (g)} WH,
        {\bf (h)} HBN, and {\bf (i)} Cam-CAN datasets.
    }
\end{figure*}

\subsubsection*{SGL accurately predicts age from tractometry data}

To test the performance of SGL with tractometry data in a continuous
regression task, we focus here on the prediction of biological age based on
tractometry data in three datasets named WH, HBN, and Cam-CAN (see
\cref{supp:sec:data} for further details). Prediction of ``brain age'' is a
commonly undertaken task in neuroimaging machine learning, in part because
these predictions, and deviations therefrom, may be diagnostic of overall
brain health (for a review, see \textcite{Cole2019-rz}). However, as
\textcite{nelson2019biomarkers} have observed, aging biomarkers are subject
to unique challenges and tend to be noncausative. Our interest in aging here
relies on its utility as a methodological benthmark. Biological age operates
on a natural scale, with meaningful and easily understood units, and it's
popularity as a machine learning target makes it valuable for comparisons
with other studies.

The WH \cite{yeatman2014lifespan}, HBN \cite{alexander2017open}, and Cam-CAN
\cite{shafto2014cambridge,taylor2017cambridge} datasets used here contain
data from 76, 978, and 640 subjects, respectively, ranging from 6-50, 5-21,
and 18-88 years of age, respectively. In each case, biological age was used
as the predicted variable ($y$ in \cref{supp:eq:lm}). SGL was fit to
tractometry-extracted features in 18 major brain tracts: DTI FA and DTI MD
for WH and DKI FA and DKI MD for HBN and Cam-CAN, with each tract divided
into 100 nodes.

To evaluate the fit of the model, we used a nested cross-validation
procedure. In this procedure, batches of subjects are held out. For each
batch (or fold), the model is fully fit without this data. Then, once the
parameters are fixed, the model is inverted to predict the ages of held out
subjects based on the linear coeffiecients. This scheme automatically finds
the right level of regularization (i.e., sparseness) and fits the
coefficients to the ill-posed linear model, while guarding against
overfitting. SGL accurately predicts the age of the subjects in this
procedure, with a median absolute error of 2.8 years, 1.5 years, and 6.0
years for WH, HBN, and Cam-CAN, respectively and coefficients of
determination $R^2 = $ {\whRsq} , {\hbnRsq}, and {\camcanRsq}, respectively (see
\cref{fig:age-results}, top panels). The predictions for Cam-CAN are
competitive with a recent state of the art prediction
\cite{mcpherson2020single}, which used streamline density to estimate the
brain's structural connectivity and achieved $R^2 = 0.63$. The median
absolute errors are also lower than the results of a recent study that
predicted age in a large sample, based on diffusion MRI features
\cite{Richard2018-ux}.

In contrast to the ALS classification case, the selected $\alpha$ values
indicate high global sparsity over group sparsity, with $\alpha = $
{\whLRatio}, {\hbnLRatio}, and {\ccLRatio}, for the WH, HBN, and Cam-CAN
datasets, respectively. The model weights are distributed over many different
tracts and dMRI tissue properties (see
\cref{fig:age-results:wh-coefs3d,fig:age-results:hbn-coefs3d,fig:age-results:cc-coefs3d}
and supplemental material
\cref{supp:fig:wh-bp:md,supp:fig:wh-bp:fa,supp:fig:hbn-bp:md,supp:fig:hbn-bp:fa,supp:fig:cc-bp:md,supp:fig:cc-bp:fa}).
This demonstrates that SGL is not coerced to produce overly sparse results
when a more accurate model requires a dense selection of features.
Furthermore, inspecting the portions of bundles with larger coefficients in
\cref{fig:age-results:wh-profile,fig:age-results:hbn-profile,fig:age-results:cc-profile}
reveals that SGL selects informative regions where diffusion properties are
different between the age quintiles.
